### 2019-05-01 Paper_reorganized.Rmd
Final draft before submission

---
title             : "Quantitative involvement moderates the interplay between mothers' versus fathers' qualitative parenting behaviors and children's emotional adjustment development"
shorttitle        : "QUALITATIVE AND QUANTITATIVE INVOLVEMENT"

@@

```{r load_packages, include = FALSE}
library("papaja")
mask <- TRUE
masked <- function(text){
  ifelse(mask, "[masked]", text)
}
```

@@

Emotional adjustment, defined as the absence of internalizing symptoms and manifest signs of emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001], is an important developmental outcome, integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013; @aldaoEmotionregulationStrategiesPsychopathology2010]. Fathers have been hypothesized to play a unique role in children's emotional adjustment development, because they encourage children to challenge themselves, in play and in life [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. The classical notion of the “attachment relationship” is rooted in security and exploration (REF). Fatherhood researchers have posited that, whereas the mother-child attachment relationship primarily provides warmth and security, exploration might be somewhat more central to the father-child relationship, which can has thus been construed as an “activation relationship” [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers are thought to stimulate a broad spectrum of emotions, thereby structuring opportunities to expand children's emotional repertoire.

@@

Although it is now widely acknowledged that both the quality and quantity of parental involvement are relevant factors in child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated. This might be, in part, because research interests have historically varied across discipline lines: Developmental psychology has focused primarily on how qualitative parenting behaviors influence child development, whereas sociology and gender studies have focused more on involvement quantity, and its implications for maternal wellbeing [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study attempts to bridge these approaches by investigating both qualitative and quantitative involvement in relation to children's emotional adjustment.

@@

The standard family environment model posits that the quality of parents' involvement is a proximal mechanism, shaping children's development [@amatoParentalDivorceMarital2008]. Qualitative parenting is typically examined in terms of the constituent dimensions of the parenting styles taxonomy [@baumrindChildCarePractices1967]: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Responsiveness refers to warm and supportive parenting, which substantial evidence indicates is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Two aspects of demandingness investigated in the present study are consistency; the extent to which consequences for incursions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. The literature on links between demandingness and emotional adjustment is more limited, and mixed [@morrisRoleFamilyContext2007]; in part because some forms of demandingness help children meet societal expectations, whereas other forms are overly restrictive, and undermine children's autonomy needs [@barberCentralityControlParenting2013]. A fourth dimension, which has consistently been linked with children's emotional maladjustment, but is distinct from the parenting styles typology, is negative parenting [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. We operationalized negative parenting as one of its specific subdimensions: parental anger. The anger parents express in interactions with children might reflect faillure to regulate their own negative emotions, and models the use of anger in communication, and undermines children's emotion regulation [@scaramellaParentingChildrenDistress2008].

<!-- 
Inductive reasoning is thought to be integral to children's development of moral emotions [@HoffmanEmpathymoraldevelopment2000], and might thus play a positive role general emotional development as well.
-->

Although some research suggests that both parents' qualitative parenting behaviors influence children in similar ways [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique roles of mothers versus fathers. Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' influence [@flouriFathersBehaviorsChildren2010]. Furthermore, whether for evolutionary or cultural reasons [@bogelsDoesFatherKnow2011], mothers and fathers may tend to adopt somewhat specialized roles in terms of qualitative parenting: The aforementioned notions of "attachment" and "activation" relationships posit that mothers empathize more with their children, and  provide more warmth and support, whereas fathers activate children by encouraging them to take risks and overcome obstacles, and live up to certain standards for behavior [@majdandzicFathersChallengingParenting2014; @paquetteTheorizingFatherchildRelationship2004]. Empirical research supports these unique roles for mothers and fathers in children's emotional adjustment development: Mothers tend to encourage the expression of emotions, and respond warmly and supportively, whereas fathers structure and sanction emotion expressions to a greater extent [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, mothers' supportive behavior was found to be a more salient socialization mechanism than control for emotional adjustment in adolescence, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. This differentiated pattern was also found with regard to younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Furthermore, there is some evidence that mothers' negative parenting has a stronger negative impact upon children's emotional adjustment than fathers' negative parenting [@changHarshParentingRelation2003]; perhaps because this behavior is incongruous with mothers' prototypical role as an attachment figure. Based on these findings, we might expect emotional adjustment to be more strongly associated with maternal warmth, and with paternal demanding behaviors, such as consistency and inductive reasoning.

@@

According to the transactional view of family relationships, parents and children reciprocally influence one another [@kerrPerceivedParentingStyle2012; @amatoParentalDivorceMarital2008]. This view of the family as a dynamic system whose members reciprocally influence one another has become widely accepted [@kerrPerceivedParentingStyle2012]. Nevertheless, examining bi-directional effects is not yet standard practice in the childhood literature [@flouriFathersBehaviorsChildren2010]. Available empirical evidence suggests that children's adjustment problems elicit less warmth, more demandingness, and more negative parenting from parents [@gadeyneLongitudinalRelationsParenting2004; @nepplIntergenerationalContinuityParenting2009; @LenguaBidirectionalassociationstemperament2005]. In the adolescent literature, bi-directional effects are more commonly examined. This might be because the notion of children influencing parents is more salient in adolescence, when parent-child relationships become increasingly egalitarian. Findings from the adolescent literature are broadly in line with findings in childhood; suggesting that emotional adjustment problems predicts decreased parental support, and greater control [@OtterpohlCrossLaggedRelationsParenting2015; @VanderGiessenPerceivedAutonomySupport2014; @Wijsbroekdirectioneffectsperceived2011]. Thus, parents' responses to maladjustment appear to be a combination of punitive responses, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. Moreover, parents' responses to adjustment problems are contingent on children's age: Because younger children are perceived to be less competent than older children, mothers respond to their misbehavior more inductively, and less punitively [@Dixaffectiveorganizationparenting1991]. We might thus expect emotional adjustment to positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

@@

Methods commonly used to examine direction of effects between parents and children have been criticized for yielding biased results if there are stable between-family differences in the variables studied [@Hamakercritiquecrosslaggedpanel2015]. Recent methodological innovations provide better estimates of dynamic reciprocal processes within families, by controlling for stable between-family differences: The random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@Hamakercritiquecrosslaggedpanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@AsparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these innovative methods, but mostly in adolescent samples. One such study found that predictive effects between parenting and delinquency adjustment that were apparent using classical methods, were found to be explained by correlations between stable between-family differences in parenting and adjustment, once these components were included in the model [@KeijsersParentalmonitoringadolescent2016]. Another study examined the interplay between maternal and paternal qualitative parenting dimensions and adolescents' emotional adjustment, and similarly found that most associations manifested as between-family level correlations. Furthermore, child effects were found to substantially outnumber parenting effects [@vanlissaRoleFathersMothers2018a]. The authors suggested that this might be due to the relative equality of adolescent-parent relationships, as compared to parent-child relationships. If this interpretation is correct, we might expect to find more prominent parenting effects in the present study, with a younger sample. These preliminary findings highlight the fact that the direction of effects between parents and children, while controlling for stable between-family differences, remains an important issue to be examined. At this stage, however, it is difficult to formulate specific hypotheses about directionality, and the level at which associations between parenting and child development should occur.

@@

@@

Quantitative involvement has been researched extensively in terms of the amount of time spent caring for the child [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the the way childcare tasks are divided is more relevant for children's adjustment than the absolute amount of time invested by each parent [@KrollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, research has shown that the equity of the childcare distribution affects child development directly [@deutschPaternalParticipationChild2001], and impacts the emotional climate of the interparental relationship [@CraigDoesFatherCare2006]. When childcare is unequally distributed, women often perform a disproportionate share of the work. This strains the marital relationship [@shockleyItNotWhat2018], and the resulting stress erodes parenting quality, particularly for parents of difficult children [@SheltonInterparentalconflictnegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement, e.g., as hours spent caring for the child, incurs a potential confound: Any effect that appears to be due to fathers' greater involvement might, in fact, be due to children receiving greater a greater total amount of parental care [see @deutschPaternalParticipationChild2001]. This confound is mitigated by operationalizing quantitative involvement in terms of the childcare distribution, and controlling for total joint care. Some prior research has attempted to capture the childcare distribution directly, by asking which has most responsibility for certain tasks, on a continuum from the mother doing everything to the father doing everything [@KrollEarlyFatherInvolvement2016]. When quantitative involvement is measured in terms of absolute involvement, however, an elegant statistical solution can disentangle the childcare distribution from total joint care. The so-called "latent mean and difference" approach, which is typically used to capture the difference between consecutive longitudinal measurements [@McArdleLatentVariableModeling2009], can be utilized to re-express mothers' and fathers' absolute quantitative involvement into a "latent mean" of involvement, which represents the total joint care provided by both parents, and latent differential involvement, which represents the extent to which one parent is more involved than the other. This approach taps directly into a theoretically relevant construct, and prevents confounding. 
## Main effects of quantitative involvement

@@

In many countries, mothers continue to spend more time with their children than fathers [@hofferth28669resident], particularly when it comes to routine care tasks [@craigHowMothersFathers2011]. Fathers' involvement is on the increase, however [@YeungChildrenTimeFathers2001]. The amount of time parents spend with children is likely to predict the quality of parenting behaviors [@sabattiniRelationMothersFathers2004], through practice and experience. Indeed, more involved parents tend to experience greater parenting self-efficacy [@GialloMothersfathersinvolvement2013]. We might thus hypothesize that greater total joint care should positively predict both parents' parenting quality.

By the same logic, we might expect that, when one parent is more involved than the other, the parenting quality of the more involved parent would improve. However, an argument to the contrary can be made as well. Retrospective reports by 19-year old participants indicate that an unequal childcare distribution tends to go hand in hand with suboptimal qualitative parenting behaviors - specifically, increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden if they come down on a single parent's shoulders [@deutschPaternalParticipationChild2001]. Mothers thus benefit directly when fathers relieve some of this burden [@PleckPaternalinvolvementRevised2004]. Consequently, mothers tend to be more satisfied when fathers are more involved [@PharesFathersMothersInvolvement2009]. An unfair distribution of childcare, on the other hand, is associated with greater maternal anger and resentment [@RossGenderParenthoodAnger1996]. This strains the interparental relationship, interferes with sensitive parenting, and undermines child wellbeing [@davies2006interparental]. In other words, when care is distributed relatively equally, the entire family system will likely be more harmonious. We might thus expect greater differential involvement to aversely affect both parents' parenting quality, as well as children's emotional adjustment.

@@

Although the quantity of paternal involvement has received substantial interest in the fields of sociology and gender studies, this literature has focused primarily on its implications for mothers' time use and wellbeing. Consequences for children's emotional outcomes have received less attention to date. Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @BronfenbrennerEcologyHumanDevelopment1979]. This suggests that the total amount of care provided should be positively associated with children's emotional adjustment. Differential involvement, on the other hand, might be negatively linked to emotional adjustment: When children perceive that one of their parents is relatively uninvolved, they likely feel less supported than when both parents are equally invested in their wellbeing [@deutschPaternalParticipationChild2001].

@@

Interactions between involvement quality and quantity have largely been overlooked in the aforementioned theoretical models of father involvement; perhaps in part because these two dimensions have rarely been jointly investigated [@JaffeeLifeFatherBenefits2003]. However, It is conceivable that children's emotional adjustment might be differently impacted by paternal control if that father is generally highly involved with the day-to-day minutiae of childrearing, than when he is generally absent. Conversely, children's emotional adjustment might elicit different qualitative parenting behaviors from hands-on fathers, who are aware of the reasons for their children's emotional troubles, than from out-of-touch fathers, who might perceive maladjustment as a problem behavior that should be sanctioned. To capture patterns like these, a model should account for interactions between involvement quality and quantity.

In this respect, theories of father involvement can be complemented by Feinberg's coparenting model [@FeinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the role of parent. Although the coparenting literature is typically invoked to describe childrearing in divorced families, it also holds insights for intact families. One dimension of coparenting is involvement quantity, which Feinberg defines as the "division of labor" in childcare. Note that this definition parallels our operationalization of quantitative involvement. The division of childcare labor is proposed to affect children's adjustment and parents' qualitative childrearing behaviors directly, and to interact with the link between these two constructs.

There is some preliminary empirical support for interactions between involvement quality and quantity; for instance, fathers' playful and caring involvement predicted preschoolers' emotional adjustment in school, but only when accompanied by a supportive coparenting relationship [@JiaTransactionalRelationsFather2012]. Furthermore, in adolescence, the quantity of paternal education-related structure, an age-appropriate form of involvement, interacted with paternal warmth to predict both academic achievement and behavioral adjustment [@WangParentalInvolvementAfrican2014]. Applying this insight from the coparenting literature, we might thus expect parents' quantitative involvement to be directly associated with their qualitative parenting behaviors and children's emotional adjustment, and to moderate effects between qualitative parenting behaviors and emotional adjustment.

@@

There is substantial evidence that parents tend to raise boys and girls differently. Prior research has indicated that parents tend to be more controlling with girls than with boys **(e.g., Pomerantz & Ruble, 1998; Smetana & Daddis, 2002)**, and express more negative parenting with boys than with girls **(Straus and Stewart 1999)**. Another point to consider is the interplay between parent and child sex. According to social learning theory, the same-sex parent plays a stronger role in a child's development **(Bussey & Bandura, 1999)**, particularly in the socialization of sex-typed behavior, such as display rules for emotions [@Brodysocializationgenderdifferences2000]. Thus, we might expect child sex to predict the same-sex parents' qualitative involvement, and associations between parenting and emotional adjustment might be stronger for same-sex parent-child dyads. There is also evidence that parental sensitivity is greatest in mother-daughter relationships, and lowest in father-son relationships **(Lovas, 2005; Schoppe-Sullivan et al., 2006)**. These findings suggest that boys tend to receive a "tougher" upbringing than girls. In sum, it is clear that sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Specifically, we might expect to find parents, display more responsiveness, structure, and inductive reasoning with girls than with boys. Boys, on the other hand, might receive more negative parenting. Furthermore, we might expect associations between qualitative parenting behaviors and emotional adjustment to be stronger between children and the same-sex parent.

There is little evidence for sex differences in emotional adjustment in young children, as indicated by a meta-analysis which found no sex differences in the temperament dimension of negative affectivity, based primarily on infant and toddler samples [@Else-questGenderdifferencestemperament2006]. Evidence for such sex differences is sometimes found in older children [@SilkAdolescentsemotionregulation2003], but these might appear and intensify in puberty [@HillIntensificationGenderRelatedRole1983]. We consequently did not expect to find sex differences in children's average levels of emotional adjustment.

@@

## Average associations between qualitative parenting dimensions and emotional adjustment

@@

### Between-family correlations

>>>>

---
title             : "Quantitative involvement moderates associations between qualitative parenting behaviors and children's emotional adjustment development"
shorttitle        : "INVOLVEMENT AND EMOTIONAL ADJUSTMENT"

@@

```{r load_packages, include = FALSE}
library("papaja")
library(motley)
source("rfunctions.R")
load("results.RData")
load("table_coefficients.RData")
mask <- TRUE
masked <- function(text){
  ifelse(mask, "[masked]", text)
}
```

@@

Emotional adjustment is defined as the absence of internalizing symptoms and manifest signs of emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. This important developmental outcome is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013; @aldaoEmotionregulationStrategiesPsychopathology2010]. According to the fatherhood literature, emotional adjustment development is one domain in which fathers might play a unique role, because fathers encourage children to challenge themselves, in play and in life [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. The classical notion of the "attachment relationship" is rooted in a balance between security and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Fatherhood researchers have posited that, whereas the mother-child attachment relationship primarily provides warmth and security, exploration might be somewhat more central to the father-child relationship, which can has thus been construed as an "activation relationship" [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers are thought to stimulate a broad spectrum of emotions, which expands children's emotional repertoire, and engages their developing regulatory abilities.

@@

Although it is now widely acknowledged that both the quality and quantity of parental involvement are relevant factors in child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004], these dimensions are rarely jointly investigated. This might be, in part, because research interests have historically varied across discipline lines: Developmental psychology has focused primarily on how qualitative parenting behaviors influence child development, whereas sociology and gender studies have focused more on involvement quantity, and its implications for maternal well-being and gender equality [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study attempts to bridge these approaches by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

The standard family environment model posits that the quality of parents' involvement is a proximal mechanism, shaping children's development [@amatoParentalDivorceMarital2008]. Qualitative parenting is typically examined in terms of the constituent dimensions of the parenting styles taxonomy [@baumrindChildCarePractices1967]: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Responsiveness refers to warm and supportive parenting, which substantial evidence indicates is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007].

The literature on links between demandingness and emotional adjustment is more limited, and mixed, than the literature on responsiveness [@morrisRoleFamilyContext2007]. This appears to be, in part, because some forms of demandingness help children meet societal expectations, whereas other forms are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Two specific aspects of demandingness, investigated in the present study, are consistency; the extent to which consequences for incursions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency has been found to correlate strongly with emotional adjustment [@brandPerceivedParentalInconsistency1990]. Inductive reasoning is thought to promote the internalization of parental expectations, and is integral to children's development of moral emotions [@hoffmanEmpathyMoralDevelopment2000]. It will thus likely play a positive role in general emotional development as well.

Negative parenting is an additional qualitative parenting dimension which is regularly included in the parenting styles literature, and appears to be negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. We operationalized negative parenting as one of its specific subcomponents: parental anger. It is thought that the anger parents express in interactions with children reflects a faillure to regulate their own negative emotions, models children's use of anger in communication, and undermines children's emotion regulation [@scaramellaParentingChildrenDistress2008].

Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' influence [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents' qualitative parenting behaviors influence children in similar ways [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique roles of mothers versus fathers. Whether for evolutionary or cultural reasons [@bogelsDoesFatherKnow2011], mothers and fathers sometimes appear to fulfill somewhat specialized roles in terms of qualitative parenting: The aforementioned notions of "attachment" and "activation" relationships posit that mothers empathize more with their children, and  provide more warmth and support, whereas fathers activate children by encouraging them to take risks and overcome obstacles, and live up to certain standards for behavior [@majdandzicFathersChallengingParenting2014; @paquetteTheorizingFatherchildRelationship2004]. Empirical research supports these unique roles for mothers and fathers in children's emotional adjustment development: Mothers tend to encourage the expression of emotions, and respond warmly and supportively, whereas fathers structure and sanction emotion expressions to a greater extent [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, mothers' supportive behavior was found to be a more salient socialization mechanism than control for emotional adjustment in adolescence, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. This differentiated pattern was also found with regard to younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Furthermore, there is some evidence that mothers' negative parenting has a stronger negative impact upon children's emotional adjustment than fathers' negative parenting [@changHarshParentingRelation2003]; perhaps because this behavior is incongruous with mothers' prototypical role as an attachment figure. Based on these findings, we might expect emotional adjustment to be more strongly associated with maternal warmth, and with paternal demanding behaviors, such as consistency and inductive reasoning.

@@

According to the transactional view of family relationships, parents and children reciprocally influence one another [@kerrPerceivedParentingStyle2012; @amatoParentalDivorceMarital2008]. This view of the family as a dynamic system whose members reciprocally influence one another has become widely accepted [@kerrPerceivedParentingStyle2012]. Nevertheless, examining bi-directional effects is not yet standard practice in the childhood literature [@flouriFathersBehaviorsChildren2010]. Available empirical evidence suggests that children's adjustment problems elicit less warmth, more demandingness, and more negative parenting from parents [@gadeyneLongitudinalRelationsParenting2004; @nepplIntergenerationalContinuityParenting2009; @lenguaBidirectionalAssociationsTemperament2005]. In the adolescent literature, bi-directional effects are more commonly examined. This might be because the notion of children influencing parents is more salient in adolescence, when parent-child relationships become increasingly egalitarian. Findings from the adolescent literature are broadly in line with findings in childhood; suggesting that emotional adjustment problems predicts decreased parental support, and greater control [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014; @wijsbroekDirectionEffectsPerceived2011]. Thus, parents' responses to maladjustment appear to be a combination of punitive responses, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. Moreover, parents' responses to adjustment problems are contingent on children's age: Because younger children are perceived to be less competent than older children, mothers respond to their misbehavior more inductively, and less punitively [@dixAffectiveOrganizationParenting1991]. We might thus expect emotional adjustment to positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

@@

Methods commonly used to examine direction of effects between parents and children have been criticized for yielding biased results if there are stable between-family differences in the variables studied [@hamakerCritiqueCrosslaggedPanel2015]. Recent methodological innovations provide better estimates of dynamic reciprocal processes within families, by controlling for stable between-family differences: The random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@asparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these innovative methods, but mostly in adolescent samples. One such study found that predictive effects between parenting and delinquency adjustment that were apparent using classical methods, were found to be explained by correlations between stable between-family differences in parenting and adjustment, once these components were included in the model [@keijsersParentalMonitoringAdolescent2016]. Another study examined the interplay between maternal and paternal qualitative parenting dimensions and adolescents' emotional adjustment, and similarly found that most associations manifested as between-family level correlations. Furthermore, child effects were found to substantially outnumber parenting effects [@vanlissaRoleFathersMothers2018a]. The authors suggested that this might be due to the relative equality of adolescent-parent relationships, as compared to parent-child relationships. If this interpretation is correct, we might expect to find more prominent parenting effects in the present study, with a younger sample. These preliminary findings highlight the fact that the direction of effects between parents and children, while controlling for stable between-family differences, remains an important issue to be examined. At this stage, however, it is difficult to formulate specific hypotheses about directionality, and the level at which associations between parenting and child development should occur.

@@

<!--
Better quality parental relationships are usually associated with more involved fathering (Belsky 1984;
Bouchard & Lee 2000; Coiro & Emery 1998; McBride & Mills 1993; Verhoeven et al. 2007). Further, when
relationships are better, this is also evident in more positive or sensitive parenting styles and more cooperative
and supportive co‑parenting (Floyd, Gilliom & Costigan 1998; Verhoeven et al. 2007).
-->

@@

Quantitative involvement has been researched extensively in terms of the absolute amount of time parents spend, or tasks they perform, in caring for the child [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the the way childcare tasks are divided is more relevant for children's adjustment than the absolute amount of time invested by each parent [@krollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, research has shown that the equity of the childcare distribution affects child development directly [@deutschPaternalParticipationChild2001], and impacts the emotional climate of the interparental relationship [@craigDoesFatherCare2006]. When childcare is unequally distributed, women often perform a disproportionate share of the work. This strains the marital relationship [@shockleyItNotWhat2018], and the resulting stress erodes parenting quality, particularly for parents of difficult children [@sheltonInterparentalConflictNegative2008]. Thus, operationalizing quantitative involvement in terms of relative involvement might capture patterns of greater relevance.

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement incurs a potential confound: Any effect that appears to be due to fathers' greater involvement might, in fact, be due to children receiving greater a greater total amount of parental care [see @deutschPaternalParticipationChild2001]. This risk is exacerbated by the fact that, within families, fathers' absolute involvement is strongly associated with mothers' absolute involvement [@russellCoparentingEarlySchoolage1994]. Operationalizing quantitative involvement in terms of the childcare distribution, and controlling for total joint care, mitigates this confound. Some prior research has attempted to capture the childcare distribution directly, by asking which parent is most responsible for certain tasks, on a continuum from the mother doing everything to the father doing everything [@krollEarlyFatherInvolvement2016]. When quantitative involvement has been measured in terms of absolute involvement, however, an elegant statistical solution can be applied to disentangle the childcare distribution from total joint care. The so-called "latent mean and difference" approach, which is typically used to capture the difference between consecutive longitudinal measurements [@mcardleLatentVariableModeling2009], can be utilized to re-express mothers' and fathers' absolute quantitative involvement into a "latent mean" of involvement, which represents the total joint care provided by both parents, and latent differential involvement, which represents the extent to which one parent is more involved than the other. This approach taps directly into a theoretically relevant construct, and prevents confounding.

A final consideration in the operationalization of quantitative involvement is that the type of tasks parents perform in the context of child care change rapidly as children develop [@wangATUSFathersInvolvement2009]. By the time children are school-aged, their dependence on instrumental care might be relatively minor. When the behaviors constituting quantitative involvement change over time, it is not clear how one might establish measurement invariance on a macro-longitudinal time scale. Consequently, in the present study, we took quantitative involvement at the first measurement wave as an indicator of between-individual differences in quantitative involvement. This means that we cannot investigate within-family dynamics pertaining to change in quantitative involvement.

## Main effects of quantitative involvement

@@

In many countries, mothers continue to spend more time with their children than fathers [@hofferth28669resident], particularly when it comes to routine care tasks [@craigHowMothersFathers2011]. Fathers' involvement is on the increase, however [@yeungChildrenTimeFathers2001]. The amount of time parents spend with children is likely to predict the quality of parenting behaviors [@sabattiniRelationMothersFathers2004], through practice and experience. Indeed, more involved parents tend to experience greater parenting self-efficacy [@gialloMothersFathersInvolvement2013a]. We might thus hypothesize that greater total joint care should positively predict both parents' parenting quality. This hypothesis is also supported by the notion of parental "buffering", which holds that, when both parents are more involved, they monitor each other's parenting behaviors, thereby ensuring that children are raised well [@mclanahanConsequencesFatherAbsence2014].

The notion that parenting quality improves through practice [@sabattiniRelationMothersFathers2004] might suggests that, when one parent is more involved than the other, the more involved parent would display more desirable qualitative parenting behaviors. However, an argument to the contrary can be made as well. The notion of parental buffering would suggest that both parents' qualitative parenting behaviors might suffer when parents are not equally involved [@mclanahanConsequencesFatherAbsence2014]. Indeed, retrospective reports by 19-year old participants indicate that an unequal childcare distribution tends to go hand in hand with suboptimal parenting behaviors: Increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden if they come down on a single parent's shoulders [@deutschPaternalParticipationChild2001]. Mothers thus benefit directly when fathers relieve some of this burden [@pleckPaternalInvolvementRevised2004]. Consequently, mothers tend to be more satisfied when fathers are more involved [@pharesFathersMothersInvolvement2009]. Moreover, balanced co-parenting appears to be associated with more desirable qualitative parenting behaviors; at least in terms of parental warmth [@bondsExaminingProcessWhich2007; @barnettInterdependenceParentingMothers2008]. An unfair distribution of childcare, on the other hand, is associated with greater maternal anger and resentment [@rossGenderParenthoodAnger1996]. This strains the interparental relationship, interferes with sensitive parenting, and undermines child wellbeing [@davies2006interparental]. In other words, when care is distributed relatively equally, the entire family system will likely be more harmonious. We might thus expect greater differential involvement to aversely affect both parents' parenting quality, as well as children's emotional adjustment.

@@

Although the quantity of paternal involvement has received substantial interest in the fields of sociology and gender studies, this literature has focused primarily on its implications for mothers' time use and wellbeing. Consequences for children's emotional outcomes have received less attention to date. Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. This suggests that the total amount of care provided should be positively associated with children's emotional adjustment. Differential involvement, on the other hand, might be negatively linked to emotional adjustment: When children perceive that one of their parents is relatively uninvolved, they likely feel less supported than when both parents are equally invested in their wellbeing [@deutschPaternalParticipationChild2001].

@@

Interactions between involvement quality and quantity have largely been overlooked in the aforementioned theoretical models of father involvement; perhaps in part because these two dimensions have rarely been jointly investigated [@jaffeeLifeFatherBenefits2003]. Children's emotional adjustment might be differently impacted by paternal control if that father is generally highly involved with the day-to-day minutiae of childrearing, than when he is generally absent. Conversely, children's emotional adjustment might elicit different qualitative parenting behaviors from hands-on fathers, who are aware of the reasons for their children's emotional troubles, than from out-of-touch fathers, who might perceive maladjustment as a problem behavior that should be sanctioned. To capture patterns like these, a model should account for interactions between involvement quality and quantity.

In this respect, theories of father involvement can be complemented by Feinberg's coparenting model [@feinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the role of parent. Although the coparenting literature is typically invoked to describe childrearing in divorced families, it also holds insights for intact families. One dimension of coparenting is involvement quantity, which Feinberg defines as the "division of labor" in childcare. Note that this definition parallels our operationalization of quantitative involvement. The division of childcare labor is proposed to affect children's adjustment and parents' qualitative childrearing behaviors directly, and to interact with the link between these two constructs.

There is some preliminary empirical support for interactions between involvement quality and quantity; for instance, fathers' playful and caring involvement predicted preschoolers' emotional adjustment in school, but only when accompanied by a supportive coparenting relationship [@jiaTransactionalRelationsFather2012]. Furthermore, in adolescence, the quantity of paternal education-related structure, an age-appropriate form of involvement, interacted with paternal warmth to predict both academic achievement and behavioral adjustment [@wangParentalInvolvementAfrican2014]. Applying this insight from the coparenting literature, we might thus expect parents' quantitative involvement to be directly associated with their qualitative parenting behaviors and children's emotional adjustment, and to moderate effects between qualitative parenting behaviors and emotional adjustment.

@@

There is substantial evidence that parents tend to raise boys and girls differently. Prior research has indicated that parents tend to be more controlling with girls than with boys [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and express more negative parenting with boys than with girls [@strausCorporalPunishmentAmerican1999]. Another point to consider is the interplay between parent and child sex. According to social learning theory, the same-sex parent plays a stronger role in a child's development [@busseySocialCognitiveTheory1999], particularly in the socialization of sex-typed behavior, such as display rules for emotions [@brodySocializationGenderDifferences2000]. Thus, we might expect child sex to predict the same-sex parents' qualitative involvement, and associations between parenting and emotional adjustment might be stronger for same-sex parent-child dyads. There is also evidence that parental sensitivity is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. These findings suggest that boys tend to receive a "tougher" upbringing than girls. In sum, it is clear that sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Specifically, we might expect to find parents, display more responsiveness, structure, and inductive reasoning with girls than with boys. Boys, on the other hand, might receive more negative parenting. Furthermore, we might expect associations between qualitative parenting behaviors and emotional adjustment to be stronger between children and the same-sex parent.

There is little evidence for sex differences in emotional adjustment in young children, as indicated by a meta-analysis which found no sex differences in the temperament dimension of negative affectivity, based primarily on infant and toddler samples [@else-questGenderDifferencesTemperament2006a]. Evidence for such sex differences is sometimes found in older children [@silkAdolescentsEmotionRegulation2003], but these might appear and intensify in puberty [@hillIntensificationGenderRelatedRole1983]. We consequently did not expect to find sex differences in children's average levels of emotional adjustment.

@@

## Associations between qualitative parenting dimensions and emotional adjustment

@@

### Between-family level correlations
